Joint maximum-likelihood of phylogeny and ancestral states is not consistent

### 2019-03-24 versions/response.v2.txt

Thank you for handling the manuscript, and please thank the reviewers for their helpful comments. As stated by Reviewer 3, we believe that the review process has significantly improved the manuscript.

> Reviewers' comments:
> Reviewer: 1

> Comments to the Author
> As with the previous version, I believe the topic matter will be of
> interest. For me, however, the work is incomplete in important ways
> that I believe can be relatively easily addressed. The authors went
> beyond my expectations in calculating explicit solutions for the
> limiting edge-lengths (Fig S2) when tau1 is the putative topology of
> estimation. This allows them to draw conclusions about edge-length
> inconsistency but there is no definitive conclusion about `strict'
> topological inconsistency (in the sense of converging upon the wrong
> tree rather than a multifurcating one). From the statements near l238
> it seems the authors have completed some preliminary investigation on
> strict topological inconsistency and find no evidence for it at
> present. Assuming that remains true, the findings of the current ms
> are of a different nature than the previous version where strict
> topological inconsistency was found. In settings with a limiting
> multifurcating topology, because it is multifurcating, for finite but
> large sequence-lengths, measures of uncertainty (eg. bootstrap
> support) will suggest that there is substantial uncertainty in
> whatever topology gets estimated (this probably deserves discussion).
> The joint optimization approach thus lacks the power to estimate the
> correct topology where ML would but is not being positively mislead.
> Assuming the lack of strict topological inconsistency holds, the
> results for the method are not as negative as they were in the
> previous version.

DISCUSS I honestly can't figure out what the reviewer wants to see re the bootstrap. Branch length inferences don't come into the bootstrap unless you contract things with zero branch lengths, which isn't generally done when doing bootstrap estimation.
das: I think the reviewer is saying that, since in the case of infinite data we converge on a multifurcating topology, that, for finite sequence lengths, we *might* estimate the incorrect topology to be of higher likelihood. The bootstrap support would just be one way to estimate how frequently this happens for different sequence lengths. We're only interested in the case of infinite data, though, so I don't think there's any benefit to this---it would just be more speculation, which we below say we don't want to do.

@@

The reviewer is correct that this story is different than the one in the previous version of the paper.

DISCUSS Shall we strike the speculation about higher likelihoods on the wrong topology?
das: I am fine with striking the speculation (and have done so, so let me know what you think). I guess I agree the results are "not as negative," though I don't think they are positive. Strong topological inconsistency would have been dope, but I don't see any way we can get it out of our method, so we'll have to settle with regular old branch length inconsistency.

@@

> Getting complete results about topological consistency is feasible.
> The arguments that led to the closed form solution for the limiting
> edge-length estimates for tau_1 can be repeated similarly for tau2.
> This would allow, for any given x* & y*, comparison of the limiting
> lnl per site for tau1 vs tau2. Whichever topology has the larger lnl
> is the long-run estimated topology. In cases where hat w=1 for both
> topologies, conclusions are different but definite. Thus it seems a
> relatively straightforward exercise given the work the authors have
> already done (I hope I am not missing a subtle difficulty) to have a
> computer determine for a grid of (x*,y*) whether (a) hat w=1 exactly
> for both topologies, (b) the difference in limiting lnl over
> topologies is either positive or negative or (c) hat w neq 1 for at
> least one of the topologies and the difference in limiting lnl is 0 up
> to machine precision. I would emphasize that the cases (a) and (b)
> constitute rigorous proof. The cases (c) likely are rare and
> correspond to boundary cases, so the boundary may have some slight
> uncertainty. There may be some aesthetic appeal to getting analytical
> results rather than using a computer but being able to draw clear
> conclusions about strict topological inconsistency is important.

DISCUSS: we have no evidence of topological inconsistency
das: I still don't quite understand what this reviewer is saying: w=1 yields the highest likelihood, so the difference in lnl will always be zero. There is no "machine precision" involved in any of it---w is one exactly. That's its optimal value. Should we change some of the discussion to hammer this home?

@@

> The language used to refer to the approach and references to the
> literature are awkward. The references to profile likelihood work
> suggest that there is a theoretical basis for the approach (that paper
> suggests consistency) but the settings considered in such work are
> quite different, I believe. Integrated likelihood usually refers to
> something different too. Finally, as Reviewer 3 makes clear the only
> current implementation of the approach was intended as an
> approximation to ML rather than as ML. The central issue is whether
> treating unobserved data (ancestral states here) as parameters
> sometimes works; the EM algorithm were developed with the idea that
> this is not what should be done. In any case, references directly tied
> to that issue would be more relevant. Additional suggestions include:
> 1) Make a clear statement in the intro about why the usual likelihood
> is considered the right likelihood: it is the probability of the
> *observed* data. 2) The point that large numbers of parameters need
> not defeat likelihood methods is valid but it should be made clear
> when referencing such work that the setting is different.

We use "integrated likelihood" in the same manner as Goldman.
We have removed references that are unrelated to the approaches considered in the manuscript, and defer these to the discussion for the interested reader to follow up with if they so choose.
We have made various terminology changes throughout.
Thank you for your suggestion in point 1; we have added it to the introduction.
das: @EM I think your discussion with Richard over an approximation being consistently biased might be helpful to answer some of the above.

@@

> -l21: Some rewording is needed. The joint method will, as n gets
> large, continue to estimate a well-resolved topology with positive
> probability; albeit with small middle edge-length.

It will estimate a branch length of zero which is equivalent to a multifurcating topology.

> -l131: A reader might be able to avoid the appendix (which is
> difficult to wade into) if an example zeta in a four taxon case was
> given as illustration.

DISCUSS eliminating xi
das: I think we should, as you're correct that we marginalize over it.
I redefined a version of it at the beginning of the proof.

@@

> -l189: `depends on the values of other branch parameters' or some
> other rewording. I don't think it is that they are unknown. It is that
> they need optimization, so you can't optimize w independently of them.

We have adopted the suggested wording, thanks.

> -l201, Figs 2 and 3: I don't understand why direct numerical
> optimization was needed. You have the exact expressions for hat w in
> Fig S2. So you could have plotted them as Fig 2. A similar comment
> applies to Fig 3. Are you trying to convince the reader that your
> proofs have no errors? That seems more an issue for supp material.

We have rewritten this section to clarify what is being used where.
%EM das-- is this right?
Specifically, although we can find the regions for which the various ancestral state splits are optimal, we cannot obtain optimal branch lengths analytically.
das: Yeah, the only "numerical optimization" is for the marginal likelihood. What's numerical in our setup is determining the regions of maximal ancestral states, and the only reason it's numerical is that analytically it would requires 72 (maybe fewer?) inequalities, so getting them "numerically" was easier.

@@

> -l252-259: I don't think this is an example of profile likelihood as
> considered in Murphy and Van der Vaart.

We agree, and have made the requisite changes.

> -l264: Do you mean that `consistency can hold even when the numbers of
> parameters tend to infinity'? The statement `not a necessary
> condition' just means that inconsistency can arise in different ways
> (eg. lack of identifiability).

We have removed this reference in the discussion (though we keep it in the introduction).
We use it to show a relatively well-known example where maximum likelihood fails for a similar reason as in the current case of joint inference of phylogeny.
The discussion preceding this sentence shows that parameters tending to infinity are not a sufficient condition for inconsistency, which we believe is more clear having removed the reference to Neyman and Scott.
We have removed the reference to the paper by LeCam.
It explains various setting where maximum likelihood fails, including Neyman and Scott's example, but is not directly relevant to our argument.

@@

This is a good point! We have rephrased things, eliminating p_ytilde_j except for a parenthetical reference to the Appendix. We believe that the result is much more readable.

> page 30 line 667-8 what are S and Y and what makes a path? I gather
> from the math that you mean paths in the tree, since the sum is over
> ancestral states and there's a product of probabilities of changes on
> branches but (perhaps my lack of background) the form of (9) isn't
> obvious I think because I am not exactly clear on what the notation
> means.

We have clarified the definitions of S and Y, and added that by path we mean graph-theoretic path. We are afraid that adding more detail about the Hadamard transform is beyond the scope of this paper, though we have added a link to a scanned page of the Semple and Steel book where I (Erick) have added some details on the proofs.

das: @EM are we keeping this imgur link or is this a placeholder for a specific reference we want to make? Do you think they'd accept this?

### 2019-03-24 appendix.tex
addressing TODOs

\subsection*{Example}
We follow with an example computing these probabilities and likelihoods.
Consider the fixed, binary four-taxon tree $\tau_1$ in Fig.~\ref{fig:farris-fels-top}a.
The set of all possible character assignments is
\begin{align*}
\mathcal{P}(\{1,2,3,4\}) &= \{\emptyset, \{1,2,3,4\}, \{1\}, \{2,3,4\}, \{2\}, \{1,3,4\}, \{3\}, \{1,2,4\}, \\
                         &\qquad \{1,2\}, \{3,4\}, \{1,3\}, \{2,4\}, \{2,3\}, \{1,4\}, \{1,2,3\}, \{1,4\}\}
\end{align*}
where each set indicates the tips assigned the character $1$.
For example, $\emptyset$ is the labeling $0000$ and $\{1,3,4\}$ is the labeling $1011$.
Symmetry allows us to group adjacent pairs in $\mathcal{P}(\{1,2,3,4\})$ into equiprobable splits, letting $\siteSplitSet=\{1,2,3\}$.
The unique site splits, collapsing complements, are
\begin{align*}
    \mathcal{P}(\siteSplitSet) &= \{\emptyset, \{1\}, \{2\}, \{3\}, \{1,2\}, \{1,3\}, \{2,3\}, \{1,2,3\}\} \\
& =: \{\siteSplit_1, \ldots, \siteSplit_8\}.
\end{align*}
Since we identify character complements, we do not consider the additional splits
\begin{equation*}
\begin{split}
& \mathcal{P}(\{1,2,3,4\}) \setminus \mathcal{P}(\siteSplitSet) = \\
&\qquad \{\{1,2,3,4\}, \{2,3,4\}, \{1,3,4\}, \{1,2,4\}, \{3,4\}, \{2,4\}, \{1,4\}, \{4\}\},
\end{split}
\end{equation*}
the symmetry of the binary character model allowing us to focus only on the elements of $\mathcal{P}(\siteSplitSet)$.
This tree has two internal nodes with $\ancestralSplitSet=\{1,2\}$ and unique ancestral state splits
\[
\mathcal{P}(\ancestralSplitSet) = \{\emptyset, \{1\}, \{2\}, \{1,2\}\}.
\]
Internal node $1$ is the node connected to leaves $1$ and $3$ while internal node $2$ is connected to leaves $2$ and $4$.
The mapping from characters to splits in this case depends on the characters at the tips and the ancestral states.
For example, we take both $\patternToSplit(0000)=\emptyset$ and $\patternToSplit(1111)=\emptyset$.
Similarly, we have $\ancestralToSplit(0000, 00) = \emptyset$ and $\ancestralToSplit(1111, 11)=\emptyset$, needing to take the complement of all the characters present on the tree to identify splits.
We cannot identify complements for ancestral states We have added a sentence clarifying language:
\[
\Pr(\siteSplitRV=\siteSplit, \ancestralSplitRV=\emptyset \mid \tau, t)\neq \Pr(\siteSplitRV=\siteSplit, \ancestralSplitRV=\{1,2\} \mid \tau, t)
\]
in general.

@@

The likelihood~\eqref{eq:profile_likelihood} can be written as
\begin{align}
L_\nCols'(\tau, t; \fullAlignment) &= \max_{\fullAncestralStates} \ L_\nCols(\tau, t; \fullAlignment, \fullAncestralStates) \nonumber \\
                             &= \prod_{i=1}^{\nCols} \ \max_{\ancestralStateColumn_i} \ \Pr(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t) \nonumber \\
                             &\propto \prod_{i=1}^{\nCols} \ \max_{\ancestralStateColumn_i} \ \Pr(\siteSplitRV=\patternToSplit(\alignmentColumn_i), \ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \tau, t) \nonumber \\
                             &= \prod_{i=1}^{\nCols} \ \Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_{\siteSplit_j} \mid \tau, t) \nonumber \\
                             &= \prod_{j=1}^{\nSiteSplits} \ \left[\Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_{\siteSplit_j} \mid \tau, t)\right] ^{\nCols_j(\fullAlignment)} \label{eq:site_pattern_likelihood}
\end{align}
for $\siteSplit_j\in\mathcal{P}(\siteSplitSet)$ and some $\xi_{\siteSplit_j}\in\ancestralSplitPartition_j(\tau, t)$ with $1 \le j \le q$ where $\nCols_j(\fullAlignment)$ is the number of columns in $\fullAlignment$ that project to site split $\siteSplit_j$.

Let
\[
L_\nCols''(\tau, t; \fullAlignment) = \prod_{j=1}^{\nSiteSplits} \ \left[\Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_{\siteSplit_j} \mid \tau, t)\right] ^{\nCols_j(\fullAlignment)}
\]
be the final product in \eqref{eq:site_pattern_likelihood}.
Assume $\nCols$ observations are generated from a model with parameters $(\tau^*, t^*)$.
We have
\begin{equation*}
\frac{1}{\nCols} \log L_\nCols''(\tau, t; \fullAlignment) = \sum_{j=1}^\nSiteSplits \frac{\nCols_j(\fullAlignment)}{\nCols}\cdot  \log \Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_{\siteSplit_j} \mid \tau, t)
\end{equation*}
so that, in the $\nCols\rightarrow\infty$ limit,
\begin{equation}
\begin{split}
&    \frac{1}{\nCols} \log L_\nCols''(\tau, t; \fullAlignment) \\
&\qquad \rightarrow \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot \log \Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_{\siteSplit_j} \mid \tau, t) \label{eq:site_pattern_profile_likelihood_mean}
\end{split}
\end{equation}
because
$\frac{\nCols_j(\fullAlignment)}{\nCols}$ converges to $\Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)$.
%EM das I believe this is the definition of \ell-- this should be made more explicit.
%EM Also, in the main text we read that \ell is the limit of L' as n goes to infinity, but here it's L''.
Here $j$ indexes the site splits, of which there are a fixed number $q$.

@@

For a fixed $\tau$, we show that $\hat{t}_\nCols \rightarrow \hat{t}$ for
\[
\hat{t}_\nCols = \arg\max_{t\in\mathcal{T}} \ \frac{1}{n}\log L_\nCols'(\tau, t; \fullAlignment)
\]
and
\[
\hat{t} = \arg\max_{t\in\mathcal{T}} \ \ell_{\tau^*,t^*}(\tau, t; \boldsymbol\xi).
\]
Using the notation in Section 5.2.1 in \citet{van1998asymptotic}, we let
\[
m_t(\alignmentColumn) = \sum_{j=1}^\nSiteSplits 1\{\patternToSplit(\alignmentColumn)=\siteSplit_j\} \cdot \log \Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_{\siteSplit_j} \mid \tau, t)
\]
so that
\[
\frac{1}{n}\log L_\nCols'(\tau, t; \fullAlignment) = \frac{1}{n} \sum_{i=1}^\nCols m_t(\alignmentColumn_i)
\]
and
\[
\ell_{\tau^*,t^*}(\tau, t; \boldsymbol\xi) = E[m_t].
\]
To show $\hat{t}_\nCols \rightarrow \hat{t}$, we use Wald's consistency proof \citep[p. 48, Theorem 5.14 of ][]{van1998asymptotic}, which requires four conditions.
The first is that $\mathcal{T}$ is compact, which is obviously true.
The second is that
\[
E\left[\sup_{t\in\mathcal{T}} m_t\right] < \infty,
\]
and, since $m_t(\alignmentColumn)$ is nonpositive for all $t$ and $\alignmentColumn$, this property holds.
The remaining conditions are on the maps
\[
\alignmentColumn \mapsto \sup_t m_t(\alignmentColumn)
\]
and
\[
t \mapsto m_t(\alignmentColumn).
\]
We need the first map to be measurable, which is evident since the domain $\alphabet^\nSiteRows$ of the mapping is a finite set, and so all subsets of the domain are also finite and thus measurable.
Finally, we must have the the second mapping be upper-semicontinuous for almost all $\alignmentColumn$.
For a fixed ancestral state split $t \mapsto m_t(\alignmentColumn)$ is continuous for all $\alignmentColumn$.
If we move about in $\mathcal{T}$, a different ancestral state split becomes more likely, though when we maximize over ancestral state splits we obtain a continuous function since the maximum over continuous functions is also continuous.
This ensures the upper-semicontinuous property of this mapping, and shows $\hat{t}_\nCols \rightarrow \hat{t}$, allowing our consistency results to be proved using $\ell_{\tau^*,t^*}(\tau, t; \boldsymbol\xi)$.

This Wald-type consistency does not extend readily to convergence of topology, i.e., we cannot use the above arguments to say that, for
\[
(\hat{\tau}_\nCols, \hat{t}_\nCols) = \arg\max_{\tau, t} \ \frac{1}{n}\log L_\nCols'(\tau, t; \fullAlignment)
\]
and
\[
(\hat{\tau}, \hat{t}) = \arg\max_{\tau, t} \ \ell_{\tau^*,t^*}(\tau, t; \boldsymbol\xi)
\]
then
\[
(\hat{\tau}_\nCols, \hat{t}_\nCols) \rightarrow (\hat{\tau}, \hat{t})
\]
as we did for $\hat{t}_\nCols$ with $\tau$ fixed due to the properties of tree space \citep{Yang1994}.
For this reason we do not claim joint inference is inconsistent in estimating topology, only that there exist parameters that yield degenerate maxima when performing joint optimization.

@@

Consider the InvFels tree $\tau_1$ with arbitrary fidelities, i.e., $t=\{x_1,y_1,x_2,y_2,w\}$.
Next we show that the likelihood $\ell_{\tau_1,t}(\tau_1, t; \boldsymbol\xi)$ remains unchanged if $x_1$ and $x_2$ are exchanged or if $y_1$ and $y_2$ are.
Although this property should not be surprising due to symmetry, we write it out for completeness.
This holds for a general $t$, and thus holds setting $t=t^*$.
Using the Hadamard transform, we calculate the generating probabilities on the InvFels tree.
For site split $\emptyset$,
\begin{align*}
    \Pr(\siteSplitRV=\emptyset\mid \tau_1, t) & = \frac{1}{8} (1 + x_1x_2 +  y_1y_2 +  x_1y_1w + x_1y_2w + y_1x_2w + x_2y_2w + x_1y_1x_2y_2) \\
                                              & = \frac{1}{8} (1 + x_1x_2 +  y_1y_2 +  w[x_1y_1 + x_1y_2 + y_1x_2 + x_2y_2] + x_1y_1x_2y_2) \\
                                              & = \frac{1}{8} (1 + x_1x_2 +  y_1y_2 +  w[x_1 + x_2][y_1 + y_2] + x_1y_1x_2y_2),
\end{align*}
and this probability is unchanged when $x_1$ is exchanged with $x_2$ and $y_1$ is exchanged with $y_2$.
Similarly, for site split $\{1,3\}$,
\[
    \Pr(\siteSplitRV=\{1,3\}\mid \tau_1, t) = \frac{1}{8} (1 + x_1x_2 +  y_1y_2 -  w[x_1 + x_2][y_1 + y_2] + x_1y_1x_2y_2),
\]
which also is invariant to exchanging $x_1$ with $x_2$ and $y_1$ with $y_2$.

@@

Thus exchanging $x_1$ with $x_2$ and $y_1$ with $y_2$ does not change the value of the log-likelihood $\ell_{\tau_1,t}(\tau_1, t; \boldsymbol\xi)$.
Therefore we can reduce the number of candidate likelihoods we need to search by, without loss of generality, assuming $x_2 \ge x_1$ and $y_2 \ge y_1$, with these likelihoods given in Table~\ref{tab:likelihoods} after maximizing over ancestral state splits.

>>>>

\subsection*{Example}
We follow with an example computing these probabilities and likelihoods.
Consider the fixed, binary four-taxon tree $\tau_1$ in Fig.~\ref{fig:farris-fels-top}a.
The set of all possible character assignments is
\begin{align*}
\mathcal{P}(\{1,2,3,4\}) &= \{\emptyset, \{1,2,3,4\}, \{1\}, \{2,3,4\}, \{2\}, \{1,3,4\}, \{3\}, \{1,2,4\}, \\
                         &\qquad \{1,2\}, \{3,4\}, \{1,3\}, \{2,4\}, \{2,3\}, \{1,4\}, \{1,2,3\}, \{1,4\}\}
\end{align*}
where each set indicates the tips assigned the character $1$.
For example, $\emptyset$ is the labeling $0000$ and $\{1,3,4\}$ is the labeling $1011$.
Symmetry allows us to group adjacent pairs in $\mathcal{P}(\{1,2,3,4\})$ into equiprobable splits, letting $\siteSplitSet=\{1,2,3\}$.
The unique site splits, collapsing complements, are
\begin{align*}
    \mathcal{P}(\siteSplitSet) &= \{\emptyset, \{1\}, \{2\}, \{3\}, \{1,2\}, \{1,3\}, \{2,3\}, \{1,2,3\}\} \\
& =: \{\siteSplit_1, \ldots, \siteSplit_8\}.
\end{align*}
Since we identify character complements, we do not consider the additional splits
\begin{equation*}
\begin{split}
& \mathcal{P}(\{1,2,3,4\}) \setminus \mathcal{P}(\siteSplitSet) = \\
&\qquad \{\{1,2,3,4\}, \{2,3,4\}, \{1,3,4\}, \{1,2,4\}, \{3,4\}, \{2,4\}, \{1,4\}, \{4\}\},
\end{split}
\end{equation*}
the symmetry of the binary character model allowing us to focus only on the elements of $\mathcal{P}(\siteSplitSet)$.
This tree has two internal nodes with $\ancestralSplitSet=\{1,2\}$ and unique ancestral state splits
\[
\mathcal{P}(\ancestralSplitSet) = \{\emptyset, \{1\}, \{2\}, \{1,2\}\}.
\]
Internal node $1$ is the node connected to leaves $1$ and $3$ while internal node $2$ is connected to leaves $2$ and $4$.
The mapping from characters to splits in this case depends on the characters at the tips and the ancestral states.
For example, we take both $\patternToSplit(0000)=\emptyset$ and $\patternToSplit(1111)=\emptyset$.
Similarly, we have $\ancestralToSplit(0000, 00) = \emptyset$ and $\ancestralToSplit(1111, 11)=\emptyset$, needing to take the complement of all the characters present on the tree to identify splits.
We cannot identify complements for ancestral states in the same way as tip states since, for $\siteSplit\in\mathcal{P}(\siteSplitSet)$,
\[
\Pr(\siteSplitRV=\siteSplit, \ancestralSplitRV=\emptyset \mid \tau, t)\neq \Pr(\siteSplitRV=\siteSplit, \ancestralSplitRV=\{1,2\} \mid \tau, t)
\]
in general.

@@

The likelihood~\eqref{eq:profile_likelihood} can be written as
\begin{align}
L_\nCols'(\tau, t; \fullAlignment) &= \max_{\fullAncestralStates} \ L_\nCols(\tau, t; \fullAlignment, \fullAncestralStates) \nonumber \\
                             &= \prod_{i=1}^{\nCols} \ \max_{\ancestralStateColumn_i} \ \Pr(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t) \nonumber \\
                             &= \prod_{i=1}^{\nCols} \ \max_{\ancestralStateColumn_i} \ \frac{1}{2}\Pr(\siteSplitRV=\patternToSplit(\alignmentColumn_i), \ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \tau, t) \nonumber \\
                             &= \prod_{i=1}^{\nCols} \ \frac{1}{2}\Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_{\siteSplit_j} \mid \tau, t) \nonumber \\
                             &= \prod_{j=1}^{\nSiteSplits} \ \left[\frac{1}{2}\Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_{\siteSplit_j} \mid \tau, t)\right] ^{\nCols_j(\fullAlignment)} \nonumber\\
                             &:= \left(\frac{1}{2}\right)^n L_\nCols''(\tau, t; \fullAlignment) \label{eq:site_pattern_likelihood}
\end{align}
for $\siteSplit_j\in\mathcal{P}(\siteSplitSet)$ and some $\xi_{\siteSplit_j}\in\ancestralSplitPartition_j(\tau, t)$ with $1 \le j \le q$ where $\nCols_j(\fullAlignment)$ is the number of columns in $\fullAlignment$ that project to site split $\siteSplit_j$.

Assume $\nCols$ observations are generated from a model with parameters $(\tau^*, t^*)$.
We have
\begin{equation*}
\frac{1}{\nCols} \log L_\nCols''(\tau, t; \fullAlignment) = \sum_{j=1}^\nSiteSplits \frac{\nCols_j(\fullAlignment)}{\nCols}\cdot  \log \Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_{\siteSplit_j} \mid \tau, t)
\end{equation*}
so that, in the $\nCols\rightarrow\infty$ limit,
\begin{equation}
\begin{split}
&    \frac{1}{\nCols} \log L_\nCols''(\tau, t; \fullAlignment) \\
&\qquad \rightarrow \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot \log \Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_{\siteSplit_j} \mid \tau, t) \\
&\qquad := \ell_{\tau^*,t^*}(\tau, t) \label{eq:site_pattern_profile_likelihood_mean}
\end{split}
\end{equation}
because $\nCols_j(\fullAlignment)/\nCols$ converges to $\Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)$.
%EM das I believe this is the definition of \ell-- this should be made more explicit.
%EM Also, in the main text we read that \ell is the limit of L' as n goes to infinity, but here it's L''.
%das: L'' is the value that L' is proportional to. It does read kinda awkward, so I just explicitly added the 1/2.
Here $j$ indexes the site splits, of which there are a fixed number $q$.

@@

For a fixed $\tau$, we show that $\hat{t}_\nCols \rightarrow \hat{t}$ for
\[
\hat{t}_\nCols = \arg\max_{t\in\mathcal{T}} \ \frac{1}{n}\log L_\nCols'(\tau, t; \fullAlignment)
\]
and
\[
\hat{t} = \arg\max_{t\in\mathcal{T}} \ \ell_{\tau^*,t^*}(\tau, t).
\]
Using the notation in Section 5.2.1 in \citet{van1998asymptotic}, we let
\[
m_t(\alignmentColumn) = \sum_{j=1}^\nSiteSplits 1\{\patternToSplit(\alignmentColumn)=\siteSplit_j\} \cdot \log \Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_{\siteSplit_j} \mid \tau, t)
\]
so that
\[
\frac{1}{n}\log L_\nCols'(\tau, t; \fullAlignment) = \frac{1}{n} \sum_{i=1}^\nCols m_t(\alignmentColumn_i)
\]
and
\[
\ell_{\tau^*,t^*}(\tau, t) = E[m_t].
\]
To show $\hat{t}_\nCols \rightarrow \hat{t}$, we use Wald's consistency proof \citep[p. 48, Theorem 5.14 of ][]{van1998asymptotic}, which requires four conditions.
The first is that $\mathcal{T}$ is compact, which is obviously true.
The second is that
\[
E\left[\sup_{t\in\mathcal{T}} m_t\right] < \infty,
\]
and, since $m_t(\alignmentColumn)$ is nonpositive for all $t$ and $\alignmentColumn$, this property holds.
The remaining conditions are on the maps
\[
\alignmentColumn \mapsto \sup_t m_t(\alignmentColumn)
\]
and
\[
t \mapsto m_t(\alignmentColumn).
\]
We need the first map to be measurable, which is evident since the domain $\alphabet^\nSiteRows$ of the mapping is a finite set, and so all subsets of the domain are also finite and thus measurable.
Finally, we must have the the second mapping be upper-semicontinuous for almost all $\alignmentColumn$.
For a fixed ancestral state split $t \mapsto m_t(\alignmentColumn)$ is continuous for all $\alignmentColumn$.
If we move about in $\mathcal{T}$, a different ancestral state split becomes more likely, though when we maximize over ancestral state splits we obtain a continuous function since the maximum over continuous functions is also continuous.
This ensures the upper-semicontinuous property of this mapping, and shows $\hat{t}_\nCols \rightarrow \hat{t}$, allowing our consistency results to be proved using $\ell_{\tau^*,t^*}(\tau, t)$.

This Wald-type consistency does not extend readily to convergence of topology, i.e., we cannot use the above arguments to say that, for
\[
(\hat{\tau}_\nCols, \hat{t}_\nCols) = \arg\max_{\tau, t} \ \frac{1}{n}\log L_\nCols'(\tau, t; \fullAlignment)
\]
and
\[
(\hat{\tau}, \hat{t}) = \arg\max_{\tau, t} \ \ell_{\tau^*,t^*}(\tau, t)
\]
then
\[
(\hat{\tau}_\nCols, \hat{t}_\nCols) \rightarrow (\hat{\tau}, \hat{t})
\]
as we did for $\hat{t}_\nCols$ with $\tau$ fixed due to the properties of tree space \citep{Yang1994}.
For this reason we do not claim joint inference is inconsistent in estimating topology, only that there exist parameters that yield degenerate maxima when performing joint optimization.

@@

Consider the InvFels tree $\tau_1$ with arbitrary fidelities, i.e., $t=\{x_1,y_1,x_2,y_2,w\}$.
Next we show that the likelihood $\ell_{\tau_1,t}(\tau_1, t)$ remains unchanged if $x_1$ and $x_2$ are exchanged or if $y_1$ and $y_2$ are.
Although this property should not be surprising due to symmetry, we write it out for completeness.
This holds for a general $t$, and thus holds setting $t=t^*$.
Using the Hadamard transform, we calculate the generating probabilities on the InvFels tree.
For site split $\emptyset$,
\begin{align*}
    \Pr(\siteSplitRV=\emptyset\mid \tau_1, t) & = \frac{1}{8} (1 + x_1x_2 +  y_1y_2 +  x_1y_1w + x_1y_2w + y_1x_2w + x_2y_2w + x_1y_1x_2y_2) \\
                                              & = \frac{1}{8} (1 + x_1x_2 +  y_1y_2 +  w[x_1y_1 + x_1y_2 + y_1x_2 + x_2y_2] + x_1y_1x_2y_2) \\
                                              & = \frac{1}{8} (1 + x_1x_2 +  y_1y_2 +  w[x_1 + x_2][y_1 + y_2] + x_1y_1x_2y_2),
\end{align*}
and this probability is unchanged when $x_1$ is exchanged with $x_2$ and $y_1$ is exchanged with $y_2$.
Similarly, for site split $\{1,3\}$,
\[
    \Pr(\siteSplitRV=\{1,3\}\mid \tau_1, t) = \frac{1}{8} (1 + x_1x_2 +  y_1y_2 -  w[x_1 + x_2][y_1 + y_2] + x_1y_1x_2y_2),
\]
which also is invariant to exchanging $x_1$ with $x_2$ and $y_1$ with $y_2$.

@@

Thus exchanging $x_1$ with $x_2$ and $y_1$ with $y_2$ does not change the value of the log-likelihood $\ell_{\tau_1,t}(\tau_1, t)$.
Therefore we can reduce the number of candidate likelihoods we need to search by, without loss of generality, assuming $x_2 \ge x_1$ and $y_2 \ge y_1$, with these likelihoods given in Table~\ref{tab:likelihoods} after maximizing over ancestral state splits.
